Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-07-22 Manuscript/MPESubmission/Revision/response.tex

\item{\textcolor{blue}{\textbf{Table 1:} The wording of the caption is off - the second paragraph is awkward. 
Are these values the probabilities of overlap between the distributions of the 'best' tree versus trees from each inference method pooled across all combinations of missing data? That is what is described in methods line 397 but it doesn't say so in the caption.
Two things:\\
1)  Comparing to the tables in Appendix C, the distribution of probabilities of overlap between max. likelihood and Bayesian trees are higher in most missing data schemes in tables in the Appendices than the overall values given in Table 1. How could the max. likelihood and Bayesian trees be so different?\\
2)  Pooling across all missing data schemes seems to wash out the effect of tree-inference method. I am not sure what a better way to present these would be, but perhaps some kind of matrix where the cells are the joint probabilities of overlap of each method with each missing data category, and then the marginal probabilities would be the probabilities of overlap for each method, averaged over missing data categories, as well as the marginal probabilities of overlap for each missing data category, averaged across methods.  e.g, RF distance:\\
Missing data:   0   25  50  75  Marginal \\
BI  1   0.9 0.8 0.75    0.863 \\
ML  1   0.9 0.75    0.25    0.725 \\
Marginal    1   0.9 0.78    0.5  \\
This may not be feasible in the text due to space limitations, maybe tables like these could be added to an appendix.}}

% TG: I don't understand this comment at all. What does "The wording of the caption is off" means? And which pararagraph is he talking about?
We fixed the highlithing in bold to the classical values of 0.05 or 0.95.
We changed the caption to:
"Bhattacharyya Coefficients of the pairwise method comparisons.
Each comparisons corresponds to the comparison of the full (125) distributions between the ``best'' tree and the ``missing data'' trees for all the four methods (Maximum Likelihood; Bayesian consensus; Maximum Likelihood Bootstraps and Bayesian posterior trees) and the two Normalised Robins-Foulds (RF) and Triplets (Tr) metrics. Each line summarizes the distribution of the probability of overlap between pairs of tree inference methods. The values highlighted in bold are the extreme values of high or low probability of overlap between two methods. If two methods have a high probability of overlap, they have a similar ability to recover the ``correct" tree topology"

1) We are not entirely sure if the reviewer means (i) why the two methods display such differences ; or (ii) why, within each method, the combination of parameters displays a difference between each single parameters ($M_LM_FM_C$ vs. $M_L$, $M_F$ and $M_C$).
In the first case, we think that the difference between the methods is due to the fact that the Maximum Likelihood trees are more likely to display ``false positive'' topologies (i.e. clades that are incorrect but can be well supported; or not) as the Bayesian consensus tree, when many data is missing, are more likely to collapse nodes, resulting in trees with less resolution but more ``conserved clades'' (i.e. the clade ((a;b);c)) could be collapsed in (a;b;c) which is still a corrected representation of the relations within the clades rather than ((b;c);a)).
In the second case, the fact that the missing data parameters display better scores alone is because in practice, when this occured, the two other parameters where set a 0\%, therefore resulting in matrices with a maximum of 75\%/2 missing data for $M_L$ and $M_F$ an 75\% maximum for $M_C$.
When combined, the maximum amount of missing data can be higher, especially when each parameter is set a 75\% missing data.

2) %TG: not entirely sure what the reviewer means by marginal probabilities. I made to extra summarizing tables in the supplementaries that may be actually easier to read than most of the others so we might want to add them to the main text.

### 2015-07-22 Manuscript/MPESubmission/Revision/OnlineAppendices-LaTeXSuppFiles/AppendixC.tex
Updated revisions

%Packages
\usepackage{pdflscape}
\usepackage{fixltx2e}
\usepackage{textcomp}
\usepackage{fullpage}
\usepackage{float}
\usepackage{latexsym}
\usepackage{url}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{array}
\usepackage[version=3]{mhchem}
\usepackage{ifthen}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{amstext}
\usepackage{enumerate}
\usepackage[osf]{mathpazo}
\usepackage{dcolumn}
\usepackage{lineno}
\pagenumbering{arabic}

@@

%Summary of the BC between pairs of methods for the MC parameter
\begin{landscape}
\begin{table}[ht]
\caption{Bhattacharyya Coefficients of the pairwise method comparisons, each of which corresponds to the normalised metric between the "best" tree and the "missing data" using either the Normalised Robinson-Foulds metric (RF) or the Normalised Triplets metric (Tr) for the $M_{C}$ missing data parameter only.}
\label{Tab_Supp_summary_BC_MC}
\centering
\begin{tabular}{lccccccc}
  \hline
 Comparison &  Metric & Min. & 1st Qu. & Median & Mean & 3rd Qu. & Max. \\  
  \hline
    Maximum Likelihood \textit{vs.} Bayesian consensus                 & $RF$ & 0.03 & 0.32 & 0.66 & 0.55 & 0.75 & 1.00 \\ 
                                                                       & $Tr$ & 0.51 & 0.69 & 0.80 & 0.76 & 0.80 & 1.00 \\ 
    Maximum Likelihood \textit{vs.} Maximum Likelihood bootstraps      & $RF$ & 0.03 & 0.17 & 0.21 & 0.31 & 0.46 & 0.68 \\ 
                                                                       & $Tr$ & 0.08 & 0.31 & 0.39 & 0.39 & 0.56 & 0.61 \\ 
    Maximum Likelihood \textit{vs.} Bayesian posterior trees           & $RF$ & 0.02 & 0.44 & 0.47 & 0.52 & 0.78 & 0.90 \\ 
                                                                       & $Tr$ & 0.21 & 0.52 & 0.59 & 0.55 & 0.66 & 0.77 \\ 
    Bayesian consensus \textit{vs.} Maximum Likelihood bootstraps      & $RF$ & 0.00 & 0.01 & 0.01 & 0.02 & 0.02 & 0.03 \\ 
                                                                       & $Tr$ & 0.08 & 0.47 & 0.62 & 0.51 & 0.66 & 0.73 \\ 
    Bayesian consensus \textit{vs.} Bayesian posterior trees           & $RF$ & 0.00 & 0.02 & 0.04 & 0.04 & 0.05 & 0.06 \\ 
                                                                       & $Tr$ & 0.21 & 0.45 & 0.64 & 0.57 & 0.74 & 0.79 \\ 
    Bayesian posterior tree \textit{vs.} Maximum Likelihood bootstraps & $RF$ & 0.69 & 0.73 & 0.73 & 0.76 & 0.81 & 0.86 \\ 
                                                                       & $Tr$ & 0.91 & 0.92 & 0.93 & 0.94 & 0.96 & 0.99 \\ 
   \hline
\end{tabular}
\end{table}
\end{landscape}
%END
\end{document}

>>>>

%Packages
\usepackage{pdflscape}
\usepackage{fixltx2e}
\usepackage{textcomp}
\usepackage{fullpage}
\usepackage{float}
\usepackage{latexsym}
\usepackage{url}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{array}
\usepackage[version=3]{mhchem}
\usepackage{ifthen}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{amstext}
\usepackage{enumerate}
\usepackage[osf]{mathpazo}
\usepackage{dcolumn}
\usepackage{lineno}
\usepackage{longtable}
\pagenumbering{arabic}

@@

%Summary of the BC between pairs of methods for the MC parameter
\begin{landscape}
\begin{table}[ht]
\caption{Bhattacharyya Coefficients of the pairwise method comparisons, each of which corresponds to the normalised metric between the "best" tree and the "missing data" using either the Normalised Robinson-Foulds metric (RF) or the Normalised Triplets metric (Tr) for the $M_{C}$ missing data parameter only.}
\label{Tab_Supp_summary_BC_MC}
\centering
\begin{tabular}{lccccccc}
  \hline
 Comparison &  Metric & Min. & 1st Qu. & Median & Mean & 3rd Qu. & Max. \\  
  \hline
    Maximum Likelihood \textit{vs.} Bayesian consensus                 & $RF$ & 0.03 & 0.32 & 0.66 & 0.55 & 0.75 & 1.00 \\ 
                                                                       & $Tr$ & 0.51 & 0.69 & 0.80 & 0.76 & 0.80 & 1.00 \\ 
    Maximum Likelihood \textit{vs.} Maximum Likelihood bootstraps      & $RF$ & 0.03 & 0.17 & 0.21 & 0.31 & 0.46 & 0.68 \\ 
                                                                       & $Tr$ & 0.08 & 0.31 & 0.39 & 0.39 & 0.56 & 0.61 \\ 
    Maximum Likelihood \textit{vs.} Bayesian posterior trees           & $RF$ & 0.02 & 0.44 & 0.47 & 0.52 & 0.78 & 0.90 \\ 
                                                                       & $Tr$ & 0.21 & 0.52 & 0.59 & 0.55 & 0.66 & 0.77 \\ 
    Bayesian consensus \textit{vs.} Maximum Likelihood bootstraps      & $RF$ & 0.00 & 0.01 & 0.01 & 0.02 & 0.02 & 0.03 \\ 
                                                                       & $Tr$ & 0.08 & 0.47 & 0.62 & 0.51 & 0.66 & 0.73 \\ 
    Bayesian consensus \textit{vs.} Bayesian posterior trees           & $RF$ & 0.00 & 0.02 & 0.04 & 0.04 & 0.05 & 0.06 \\ 
                                                                       & $Tr$ & 0.21 & 0.45 & 0.64 & 0.57 & 0.74 & 0.79 \\ 
    Bayesian posterior tree \textit{vs.} Maximum Likelihood bootstraps & $RF$ & 0.69 & 0.73 & 0.73 & 0.76 & 0.81 & 0.86 \\ 
                                                                       & $Tr$ & 0.91 & 0.92 & 0.93 & 0.94 & 0.96 & 0.99 \\ 
   \hline
\end{tabular}
\end{table}
\end{landscape}

\begin{table}[ht]
\caption{Summary of the median Bhattacharyya Coefficients comparisons between the missing data parameters. Each value corresponds to the median Bhattacharyya Coefficient between the Normalised metrics (Robinson-Foulds - RF; Triplets - Tr) between the ``best'' and the ``missing data'' trees (as outlined in Figure 2 in the main text) for the the trees with: Missing Living taxa only ($M_L$); Missing Fossil data only ($M_F$); Missing characters only ($M_C$); and the combination of the three parameters ($M_LM_FM_C$).} 
\begin{tabular}{llcccc}  
  \hline
 Method & Metric & $M_L$ & $M_F$ & $M_C$ & $M_LM_FM_C$ \\ 
  \hline
Maximum Likelihood & RF & 0.54 & 0.37 & 0.56 & 0.54 \\ 
                   & Tr & 0.80 & 0.75 & 0.80 & 0.79 \\ 
Bayesian consensus & RF & 0.73 & 0.59 & 0.60 & 0.79 \\ 
   & Tr & 0.86 & 0.67 & 0.81 & 0.81 \\ 
Maximum Likelihood bootstraps & RF & 0.98 & 0.96 & 0.97 & 0.78 \\ 
   & Tr & 0.99 & 1.00 & 1.00 & 0.96 \\ 
Bayesian posterior trees & RF & 0.86 & 0.93 & 0.94 & 0.68 \\ 
  & Tr & 0.98 & 0.99 & 0.99 & 0.91 \\ 
 \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{Summary of the median Bhattacharyya Coefficients comparisons between the missing data parameters and the methods. Each value corresponds to the median Bhattacharyya Coefficient between the Normalised metrics (Robinson-Foulds - RF; Triplets - Tr) between the `methods (as outlined in Figure 3 in the main text) for the the trees with: Missing Living taxa only ($M_L$); Missing Fossil data only ($M_F$); Missing characters only ($M_C$); and the combination of the three parameters ($M_LM_FM_C$).}
\begin{tabular}{llcccc}  
  \hline
 Method & Metric & $M_L$ & $M_F$ & $M_C$ & $M_LM_FM_C$ \\ 
  \hline
Maximum likelihood vs. Bayesian consensus & RF & 0.69 & 0.48 & 0.66 & 0.10 \\ 
   & Tr & 0.84 & 0.75 & 0.80 & 0.61 \\ 
Maximum likelihood vs. Bootstraps & RF & 0.29 & 0.32 & 0.21 & 0.69 \\ 
   & Tr & 0.53 & 0.40 & 0.39 & 0.65 \\ 
Maximum likelihood vs. Bayesian posterior trees & RF & 0.61 & 0.71 & 0.47 & 0.80 \\ 
  & Tr & 0.70 & 0.62 & 0.59 & 0.73 \\ 
Bayesian consensus vs. Bootstraps & RF & 0.02 & 0.01 & 0.01 & 0.00 \\ 
  & Tr & 0.78 & 0.54 & 0.62 & 0.59 \\ 
Bayesian consensus vs. Bayesian posterior trees & RF & 0.02 & 0.02 & 0.04 & 0.01 \\ 
   & Tr & 0.75 & 0.66 & 0.64 & 0.56 \\ 
Boostraps vs. Bayesian posterior trees & RF & 0.85 & 0.72 & 0.73 & 0.85 \\ 
   & Tr & 0.96 & 0.91 & 0.93 & 0.98 \\ 
 \hline
\end{tabular}
\end{table}
%END
\end{document}

### 2015-07-22 Manuscript/MPESubmission/Revision/TEM_revised_manuscript.tex
Updated revisions

\noindent
\textbf{Table 1:} Bhattacharyya Coefficients of the pairwise method comparisons.
Each comparison which corresponds to the normalised metric between the ``best'' tree and the ``missing data" trees using either the Normalised Robinson-Foulds metric (RF) or the Normalised Triplets metric (Tr). Each line summarizes the distribution of the probability of overlap between pairs of tree inference methods. Note that this is equivalent to performing a two-sided t-test, but we use the Bhattacharyya Coefficient here because we are comparing distributions not means. The values highlighted in bold are the extreme values of high or low probability of overlap between two methods. If two methods have a high probability of overlap, they have a similar ability to recover the ``correct" tree topology.\\

\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\caption{}
\centering
\begin{tabular}{|l|c|c|c|c|c|c|c|}
  \hline
 Comparison &  Metric & Min. & 1st Qu. & Median & Mean & 3rd Qu. & Max. \\ 
  \hline
    Maximum Likelihood \textit{vs.} Bayesian consensus                 & $RF$ & 0.00 & 0.00 & 0.10 & 0.20 & 0.32 & 1.00 \\ 
                                                                       & $Tr$ & 0.34 & 0.49 & 0.61 & 0.62 & 0.75 & 1.00 \\ 
    Maximum Likelihood \textit{vs.} Maximum Likelihood bootstraps      & $RF$ & 0.03 & 0.54 & 0.69 & 0.64 & 0.77 & 0.98 \\ 
                                                                       & $Tr$ & 0.08 & 0.57 & 0.65 & 0.64 & 0.73 & 0.82 \\ 
    Maximum Likelihood \textit{vs.} Bayesian posterior trees           & $RF$ & 0.02 & 0.74 & 0.80 & 0.79 & 0.89 & 0.98 \\ 
                                                                       & $Tr$ & 0.21 & 0.67 & 0.73 & 0.72 & 0.77 & 0.84 \\ 
    Bayesian consensus \textit{vs.} Maximum Likelihood bootstraps      & $RF$ & 0.00 & 0.00 & \textbf{0.00} & \textbf{0.01} & 0.01 & 0.04 \\ 
                                                                       & $Tr$ & 0.08 & 0.38 & 0.59 & 0.57 & 0.73 & 0.84 \\ 
    Bayesian consensus \textit{vs.} Bayesian posterior trees           & $RF$ & 0.00 & 0.00 & \textbf{0.01} & \textbf{0.02} & 0.04 & 0.11 \\ 
                                                                       & $Tr$ & 0.21 & 0.36 & 0.56 & 0.55 & 0.74 & 0.87 \\ 
    Bayesian posterior tree \textit{vs.} Maximum Likelihood bootstraps & $RF$ & 0.50 & 0.77 & \textbf{0.85} & \textbf{0.85} & 0.96 & 1.00 \\ 
                                                                       & $Tr$ & 0.91 & 0.96 & \textbf{0.98} & \textbf{0.97} & 0.99 & 1.00 \\ 
   \hline
\end{tabular}
\label{Tab_Results-Difference_methods}
%\end{adjustwidth}
\end{table}
\end{landscape}

>>>>

\noindent
\textbf{Table 1:} Bhattacharyya Coefficients of the pairwise method comparisons.
Each comparisons corresponds to the comparison of the full (125) distributions between the ``best'' tree and the ``missing data'' trees for all the four methods (Maximum Likelihood; Bayesian consensus; Maximum Likelihood Bootstraps and Bayesian posterior trees) and the two Normalised Robins-Foulds (RF) and Triplets (Tr) metrics. Each line summarizes the distribution of the probability of overlap between pairs of tree inference methods. The values highlighted in bold are the extreme values of high or low probability of overlap between two methods. If two methods have a high probability of overlap, they have a similar ability to recover the ``correct" tree topology.\\

\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\caption{}
\centering
\begin{tabular}{|l|c|c|c|c|c|c|c|}
  \hline
 Comparison &  Metric & Min. & 1st Qu. & Median & Mean & 3rd Qu. & Max. \\ 
  \hline
    Maximum Likelihood \textit{vs.} Bayesian consensus                 & $RF$ & \textbf{0.00} & \textbf{0.00} & 0.10 & 0.20 & 0.32 & \textbf{1.00} \\ 
                                                                       & $Tr$ & 0.34 & 0.49 & 0.61 & 0.62 & 0.75 & \textbf{1.00} \\ 
    Maximum Likelihood \textit{vs.} Maximum Likelihood bootstraps      & $RF$ & \textbf{0.03} & 0.54 & 0.69 & 0.64 & 0.77 & \textbf{0.98} \\ 
                                                                       & $Tr$ & 0.08 & 0.57 & 0.65 & 0.64 & 0.73 & 0.82 \\ 
    Maximum Likelihood \textit{vs.} Bayesian posterior trees           & $RF$ & \textbf{0.02} & 0.74 & 0.80 & 0.79 & 0.89 & \textbf{0.98} \\ 
                                                                       & $Tr$ & 0.21 & 0.67 & 0.73 & 0.72 & 0.77 & 0.84 \\ 
    Bayesian consensus \textit{vs.} Maximum Likelihood bootstraps      & $RF$ & \textbf{0.00} & \textbf{0.00} & \textbf{0.00} & \textbf{0.01} & \textbf{0.01} & \textbf{0.04} \\ 
                                                                       & $Tr$ & 0.08 & 0.38 & 0.59 & 0.57 & 0.73 & 0.84 \\ 
    Bayesian consensus \textit{vs.} Bayesian posterior trees           & $RF$ & \textbf{0.00} & \textbf{0.00} & \textbf{0.01} & \textbf{0.02} & \textbf{0.04} & 0.11 \\ 
                                                                       & $Tr$ & 0.21 & 0.36 & 0.56 & 0.55 & 0.74 & 0.87 \\ 
    Bayesian posterior tree \textit{vs.} Maximum Likelihood bootstraps & $RF$ & 0.50 & 0.77 & 0.85 & 0.85 & \textbf{0.96} & \textbf{1.00} \\ 
                                                                       & $Tr$ & 0.91 & \textbf{0.96} & \textbf{0.98} & \textbf{0.97} & \textbf{0.99} & \textbf{1.00} \\ 
   \hline
\end{tabular}
\label{Tab_Results-Difference_methods}
%\end{adjustwidth}
\end{table}
\end{landscape}
